\\Large Rising temperatures alter how and where mammals move

### 2025-09-29 writing/manuscript.Rmd
addressed minor edits throughout

We used the the `climatenaR` package [v. 1.0, @burnett_climatenar_2023] for `R` and ClimateNA [v. 7.4.2$;$ @wang_locally_2016; @mahony_global_2022] to obtain projected average monthly temperatures from 2025 to 2100 at a 0.08$^\circ$ spatial resolution. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the historical data for western Canada from 1998 to 2023 (extremes included) to estimate the monthly variation in temperature, which we defined as $\sigma^2_T$, the variance at location $\langle x, y \rangle$ in month $m$ for each year (e.g., the variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005). We then modeled the estimated variance using a Generalized Additive Model (GAM) for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] using the `mgcv` package for `R` [v. 1.9-1, @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of: each year's estimated within-pixel monthly mean temperature ($\hat \mu_T$), month ($m$), a two-dimensional smooth of spatial coordinates $\langle x, y \rangle$, and a tensor product interaction term of month and space to allow for seasonal terms to vary smoothly over space. The linear predictor for the scale term, which governs the mean-variance relationship [table 3.1 on p. 104 of @wood_generalized_2017], included smooth terms of the monthly mean temperature, month, and space. We did not include smooth terms of year to avoid unrealistic projections when extrapolating beyond past 2023. The model was thus

\begin{equation}\label{eq:temp-var-model}
\begin{cases}
\sigma^2_T \sim \text{Gamma}(\mu_{\sigma^2_T}, \nu_{\sigma^2_T}) \\
\mathbb E(\sigma^2_T) = \mu_{\sigma^2_T} \\
\mathbb V(\sigma^2_T) = (\mu_{\sigma^2_T})^2 / \nu_{\sigma^2_T} \\
\log(\mu_{\sigma^2}) = \beta_{L,0} + f_{L,1}(\mu_T) + f_{L,2}(m) + f_{L,3}(x, y) + f_{L,4}(x, y, m) \\
\log(\nu_{\sigma^2}) = \beta_{S,0} + f_{S,1}(\mu_T) + f_{S,2}(m) + f_{S,3}(x, y)
\end{cases},
\end{equation}

@@

### Resource rasters

@@

The two following sections detail the statistical framework and models we used to estimate the effect on temperature on the six species' movement behaviour. To assess the importance of including temperature as an explicit covariate (as ooposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017. See Appendix B for additional information.

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`, in meters), and distance from water (`dist_water_m`, in meters) by fitting a Hierarchical Resource Selection Function (HRSF) for each species [@mccabe_resource_2021]. We fit each HRSF using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008]. After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019; @alston_mitigating_2022] to produce estimates of second-order habitat selection [@johnson_comparison_1980]. Quadrature points were determined using the raster cells in each animal's 99.9% AKDE percentile and given a weight of 1. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B12), especially after accounting for the AKDE weights (Fig. B13).

@@

\noindent Species' changes in movement rates to temperature varied in both direction and magnitude (Figs. \@ref(fig:dist), B4-B6), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch. 11 of @mcelreath_statistical_2020]. All species altered their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The clearest example of this was cougars. In late spring (June 1^st^), they moved from evening to early morning if temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM. Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year. All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig. B8). Uncertainties around the estimated effects were generally higher at extreme temperatures due to lower data availability (Figs. B4A, B5A, and B6A).

@@

As temperatures warm, mammals' risk of heperthermia increases nonlinearly [@sherwood_adaptability_2010], especially for large-bodied species [@alston_temperature_2020; @dyer_travel_2023; @verzuh_behavioural_2023]. The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behaviour and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012]. A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013]. Firstly, exposure to warming temperatures is widespread throughout the area, and warming is expected to accelerate over the coming decades [@ipcc_climate_2023; @diffenbaugh_changes_2013; @bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023]. Secondly, large-bodied mammals are particularly sensitive to high heat [@dyer_travel_2023], but small-bodied mammals are also affected [@ratnayake_forecasting_2019; @krebs_impact_2019]. Finally, species' adaptability is tested by the speed of climate change [@diffenbaugh_changes_2013] and the rigidity of many species' photoperiod-dependent phenology [@walker_global_2019].

@@

Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes. Plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013]. Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023]. Additionally, plants at high elevations may be have weaker defenses against herbivory and be more susceptible to overgrazing [@callis-duehl_community-level_2017]. The scarcity and susceptibility of forage may thus limit herbivores' ability to shift elevationally, particularly if predation rates remain relatively unchanged or increase. Furthermore, the physical upper limit to elevation can cause individuals to become trapped . If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may become too excessive to move across patches or alternative ranges [@white_mountain_2025]. Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations can be highly species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].

@@

Our results demonstrate that mammals respond to changes in temperature by altering how and where they move in complex and nonlinear ways (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)). Hierarchical Generalized Additive Models were crucial in producing such results. The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously by inexperienced analysts. We suggest choosing the basis size (`k`) for each smooth term carefully. While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].

>>>>

To obtain projected average monthly temperatures from 2025 to 2100 at a 0.08$^\circ$ spatial resolution, we used the the `climatenaR` package [v. 1.0, @burnett_climatenar_2023] for `R` and ClimateNA v. 7.4.2 [@wang_locally_2016; @mahony_global_2022]. However, the climate projections only included estimates of future monthly averages, a scale substantially coarser than that of our tracking data (Table \@ref(tab:data-table)). To estimate the distribtution of temperatures at an hourly scale within a month, we assumed values to be approximately normally distributed with mean $\mu_T$ and variance $\sigma^2_T$. We estimated $\mu_T$ using the ClimateNA projections, while we estimated $\sigma^2_T$ by modeling the observed variance in historical weather data for western Canada from 1998 to 2023 (inclusive). For each location $\langle x, y \rangle$ and month $m$ of each year (e.g., the observed variance at coordinates $\langle -119.40, 49.94 \rangle$ in April 2005), we calculated the variance in historical temperature data, for a total of $(2024 - 1998) \times 12 = 312$ observations per location. We then modeled $\sigma^2_T$ a Generalized Additive Model (GAM) for Location and Scale [GAMLS, @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017] fitted with the `mgcv` package for `R` [v. 1.9-1, @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of: each year's estimated within-pixel monthly mean temperature ($\hat \mu_T$), month ($m$), a two-dimensional smooth of spatial coordinates $\langle x, y \rangle$, and a tensor product interaction term of month and space to allow for seasonal terms to vary smoothly over space. The linear predictor for the scale term, which governs the mean-variance relationship [table 3.1 on p. 104 of @wood_generalized_2017], included smooth terms of the monthly mean temperature, month, and space. We did not include smooth terms of year to avoid unrealistic projections when extrapolating beyond past 2023. The complete model for the distribution of projected temperature was thus

\begin{equation}\label{eq:temp-var-model}
\begin{cases}
T \sim \text{Normal}(\mu_T, \sigma^2_T) \\
\mu_T \approx \hat \mu_T \text{ as projected by ClimateNA} \\
\sigma^2_T \sim \text{Gamma}(\mu_{\sigma^2_T}, \nu_{\sigma^2_T}) \\
\mathbb E(\sigma^2_T) = \mu_{\sigma^2_T} \\
\mathbb V(\sigma^2_T) = (\mu_{\sigma^2_T})^2 / \nu_{\sigma^2_T} \\
\log(\mu_{\sigma^2}) = \beta_{L,0} + f_{L,1}(\mu_T) + f_{L,2}(m) + f_{L,3}(x, y) + f_{L,4}(x, y, m) \\
\log(\nu_{\sigma^2}) = \beta_{S,0} + f_{S,1}(\mu_T) + f_{S,2}(m) + f_{S,3}(x, y)
\end{cases},
\end{equation}

@@

### Habitat rasters

@@

To assess the importance of including temperature as an explicit covariate (as ooposed to including its effects with time of day and day of year), we fit models with and without smooth effects of temperature and compared the fits of the two sets of models via analyses of deviance (a form of generalized likelihood ratio tests) following the methods of Section 3.3 in @wood_generalized_2017. See Appendix B for additional information.

@@

\noindent We estimated the effects of temperature on each species' selection for percent forest cover (`forest_perc`), elevation (`elevation_m`, in meters), and distance from water (`dist_water_m`, in meters) by fitting a Hierarchical Resource Selection Function (HRSF) for each species [@mccabe_resource_2021]. We fit each HRSF using an HGAM with a Poisson family of distributions and log link function [Appendix B$;$ @aarts_estimating_2008]. After removing non-resident individuals (Table B1), we accounted for the spatiotemporal autocorrelation in the telemetry locations by weighting each point based on the telemetry's Autocorrelated Kernel Density Estimate [@fleming_new_2017; @noonan_comprehensive_2019; @alston_mitigating_2022] to produce estimates of second-order habitat selection [@johnson_comparison_1980]. Quadrature points were used to approximate the likelihood function of a Poisson point process through Monte-Carlo Markov chain-based integration [@baddeley_spatial_2015; @fithian_finite-sample_2013; @aarts_estimating_2008] and were determined using the raster cells in each animal's 99.9% AKDE percentile, with each raster cell having a weight of 1. The number of quadrature locations greatly outnumbered the number of observed locations (Fig. B12), especially after accounting for the AKDE weights (Fig. B13).

@@

\noindent Species' changes in movement rates to temperature varied in both direction and magnitude (Figs. \@ref(fig:dist), B4-B6), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses]. All species altered their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The clearest example of this was cougars. In late spring (June 1^st^), they moved from evening to early morning if temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM. Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year. All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig. B8). Uncertainties around the estimated effects were generally higher at extreme temperatures due to lower data availability (Figs. B4A, B5A, and B6A).

@@

As temperatures warm, mammals' risk of heperthermia increases nonlinearly [@sherwood_adaptability_2010], especially for large-bodied species [@alston_temperature_2020; @dyer_travel_2023; @verzuh_behavioural_2023]. The recent increase in frequency and duration of extreme-temperature events [@yao_emergence_2022; @ipcc_climate_2023; @zhang_increased_2023; @bunnell_global_2011; @kienzle_has_2018] have forced many mammals towards increasingly atypical behaviours with long-term consequences, ranging from changes in the behaviour and fitness of individuals [@brivio_seeking_2024; @cunningham_opportunity_2021] to mass die-offs [@ameca_y_juarez_natural_2012]. A species' ability to respond to changing temperature and phenology depends mainly on the its exposure, sensitivity, and adaptability to changes, and growing evidence suggests that many mammals in western Canada will struggle to respond due to limitations in all three factors [@ameca_y_juarez_natural_2012; @diffenbaugh_changes_2013]. Firstly, exposure to warming temperatures is widespread throughout the area, and warming is expected to accelerate over the coming decades [@ipcc_climate_2023; @diffenbaugh_changes_2013; @bunnell_global_2011; @kienzle_has_2018; @eum_potential_2023]. Secondly, large-bodied mammals are particularly sensitive to high heat [@dyer_travel_2023], but small-bodied mammals are also affected [@ratnayake_forecasting_2019; @krebs_impact_2019]. Finally, species' adaptability is tested by the speed of climate change [@diffenbaugh_changes_2013] and the rigidity of many species' photoperiod-dependent phenology [@walker_global_2019].

@@

Forage quality and availability is often a limiting factor for mammals' ability to adapt to warming temperatures by moving to higher altitudes. Plants are slower to arrive in and establish in new habitats than mammals, and the advance of herbivores can further limit plants' ability to establish [@speed_elevational_2012; @diffenbaugh_changes_2013]. Consequently, mammals that move to higher elevations may be forced to spend more time foraging [@mezzini_how_2025], which will increase their energetic expenditure as well as their chances of encountering predators (including cougars and grizzly bears) or competitors [@kohl_prey_2019; @martinez-garcia_how_2020; @torrez-herrera_monkeys_2020; @smith_densitydependent_2023]. Additionally, plants at high elevations may have weaker defenses against herbivory and be more susceptible to overgrazing [@callis-duehl_community-level_2017]. The scarcity and susceptibility of forage may thus limit herbivores' ability to shift elevationally, particularly if predation rates remain relatively unchanged or increase. Furthermore, the physical upper limit to elevation can cause individuals to become trapped . If local peaks become unsuitable habitat, and moving to nearby higher peaks requires traveling across low-quality and high-risk habitat (Figure \@ref(fig:s-rss-cc)), habitat fragmentation may become too excessive to move across patches or alternative ranges [@white_mountain_2025]. Translocating animals to locations with higher peaks may help avoid such a scenario, but the success of relocations can be highly species-dependent, and attempts are not always successful [@ranc_memory_2022; @white_mountain_2025].

@@

Our results demonstrate that mammals respond to changes in temperature by altering how and where they move in complex and nonlinear ways (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)). Hierarchical Generalized Additive Models were crucial in producing such results. The hierarchical approach us to produce statistically sound population-level estimates by accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @mortlock_sleep_2024; @jeltsch_need_2025; @silva_too_2025], while the flexibility of GAMs estimated mammals' changes in movement behaviour without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously. We suggest choosing the basis size (`k`) for each smooth term carefully. While penalized splines and REML can help avoid unruly smooth terms that would result in questionable estimates the penalty terms can struggle to restrict model complexity when datasets are particularly large [@wood_generalized_2017; @wood_generalized_2017-1].
